An open and continuously updated fern tree of life (FTOL)

### 2022-02-10 ms/manuscript.Rmd
Update MS

```{r load-targets}
# Targets from FTOL workflow
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		sanger_tree_fast,
		plastome_tree,
		plastid_tree_dated),
	# FIXME: change to ftol_cache when ready
	store = here::here("working/_targets")
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files
	),
	store = here::here("_targets")
)
```

@@

We dated the tree using treePL [@Smith2012].
We applied a set of FIXME calibration points as minimum ages.
Our calibration points are all based on fossils following a careful literature review.
We used the species identification of each fossil to assign lineage only after consulting the original publication; previous studies [`r eg`, @Testo2016a] have used genus to assign lineage, but this can be problematic as genus concepts may vary between the original publication and current taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
We tested rate smoothing parameters ranging from 0.000001 to 1000 (each varies by one order of magnitude); a value of FIXME was selected with the smallest chi-squared value for final analysis [@Smith2012].

@@

# Format for text in MS
tax_coverage <-
	coverage_data %>%
	filter(rank == "plot_group") %>%
	arrange(coverage) %>%
	transmute(clade = name, coverage = percent(coverage)) %>%
	split(.$clade)

# Format overall sampling for printing in MS
total_sampling <-
	# Start with FTOL sampling by genus, family, order
	ftol_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		sanger_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted),
		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
	# Split for printing
	split(.$rank)

@@

# Format sampling for printing in MS
bb_total_sampling <-
	# Start with backbone sampling by genus, family, order
	bb_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		bb_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted),
		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
	# Split for printing
	split(.$rank)
```

@@

Within Aspleniineae, we recover two large clades.
One clade consists of Cyctopteridaceae, Rhacidosoraceae, Diplaziopsidaceae, and a clade comprised of Aspleniaceae and Hemidictyaceae.
The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and a clade comprised of Athyriaceae and Desmophlebiaceae.

@@

@@

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(
		species = get_tips_in_ape_plot_order(plastome_tree_rooted)
	) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		plot_group = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ plot_group)
	)
```

@@

genbank_plot <-
	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
	geom_line(aes(color = type)) +
	scale_color_manual(values = compartment_cols, guide = "none") +
	# Annotate panel border manually, since the automatic border gets
	# printed on top of label annotations with theme_bw()
	annotate(
		"rect", 
		xmin = min(gb_species_by_year$year) - year_buffer,
		xmax = max(gb_species_by_year$year) + year_buffer,
		ymin = min(gb_species_by_year$n_species) - sp_buffer,
		ymax = max(gb_species_by_year$n_species) + sp_buffer,
		fill = "transparent",
		color = "grey20") +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("nuclear", "mitochondrial")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
		# Do not repel from top or bottom edges.
		ylim = c(-Inf, Inf)
	) +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("total", "plastid")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
		ylim = c(-Inf, Inf),
		min.segment.length = 0
	) +
	geom_point(
		data = fern_tree_sampling,
		aes(shape = source)
	) +
	scale_shape(
		limits = fern_tree_sampling$source, 
		labels = fern_tree_sampling$label
	) +
	scale_x_continuous(expand = expansion(0, 0)) + 
	scale_y_continuous(expand = expansion(0, 0)) + 
	labs(shape = "Study", y = "No. species", x = "Year") +
	coord_cartesian(clip = "off") +
	guides(
		shape = guide_legend(
			title.position = "top",
			nrow = 2)
	) +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		legend.position = "bottom",
		plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")
	)
```

@@

```{r backbone-make}
# Make exemplar set of species in backbone representing families
family_exemp <-
	backbone_taxonomy %>%
	# Slice to one exemplar species per family
	group_by(family) %>%
	slice(1) %>%
	ungroup() %>%
	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
	arrange(phy_order)

# Trim tree to exemplar species, rename by family
phy <- keep.tip(plastome_tree, family_exemp$species)
phy$tip.label <- family_exemp$family
# Convert BS support values to numeric
phy$node.label <- parse_number(phy$node.label)
# Reroot, ladderize, drop outgroup
phy <-
	phy %>%
	root.phylo("og") %>%
	ladderize() %>%
	drop.tip("og") %>%
	# Scale to total height of 100 for easy plotting
	rescale_tree(100)

@@

>>>>

```{r load-targets}
# Targets from FTOL workflow

# Specify path to FTOL cache
# FIXME: change to ftol_cache when ready
ftol_cache <- here::here("working/_targets")

tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		sanger_tree_fast,
		sanger_sampling,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files,
		other_dates
	),
	store = here::here("_targets")
)
```

@@

```{r dating-methods-stats}
# Count number of calibration points
n_calibrations_ferns <-
	fossil_calibration_tips %>%
	assert(is_uniq, node_calibrated) %>%
	filter(!affinities %in% c("Tracheophytes", "Euphyllophytes")) %>%
	pull(node_calibrated) %>%
	n_distinct() %>%
	number()

n_calibrations_total <-
	fossil_calibration_tips %>%
	assert(is_uniq, node_calibrated) %>%
	pull(node_calibrated) %>%
	n_distinct() %>%
	number()

# Format values used for treePL smoothing as tibble
cv_smooth_test <-	
	tibble(cv_result = treepl_cv_results) %>%
	mutate(smooth = str_match(cv_result, "\\((.*)\\)") %>% 
				 	magrittr::extract(, 2) %>%
				 	parse_number()) %>%
	mutate(chisq = str_match(cv_result, "\\) (.*)$") %>% 
				 	magrittr::extract(, 2) %>%
				 	parse_number()) %>%
	arrange(smooth)

# Extract start, end, and best smooth values
smooth_start <- cv_smooth_test %>%
	slice_min(n = 1, order_by = smooth) %>%
	pull(smooth) %>%
	number(., accuracy = .)

smooth_end <- cv_smooth_test %>%
	slice_max(n = 1, order_by = smooth) %>%
	pull(smooth) %>%
	number(., accuracy = .)

smooth_best <-
	cv_smooth_test %>%
	slice_min(n = 1, order_by = chisq) %>%
	pull(smooth) %>%
	number(., accuracy = .)
```

We dated the tree using treePL [@Smith2012].
We applied a set of `r n_calibrations_total` calibration points as minimum ages, including `r n_calibrations_ferns` calibration points within ferns.
Our calibration points are all based on fossils.
We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil name), since taxon concepts may vary between the original publication and currently applied taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); a value of `r smooth_best` was selected with the smallest chi-squared value for final analysis [@Smith2012].

@@

# Format for text in MS
tax_coverage <-
	coverage_data %>%
	filter(rank == "plot_group") %>%
	arrange(coverage) %>%
	transmute(clade = name, coverage = percent(coverage, accuracy = 0.1)) %>%
	split(.$clade)

# Format overall sampling for printing in MS
total_sampling <-
	# Start with FTOL sampling by genus, family, order
	ftol_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		sanger_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted, accuracy = 0.1),
		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
	# Split for printing
	split(.$rank)

@@

# Format sampling for printing in MS
bb_total_sampling <-
	# Start with backbone sampling by genus, family, order
	bb_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		bb_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted),
		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 0.1)}")) %>% #nolint
	# Split for printing
	split(.$rank)
```

@@

Within Aspleniineae, Cystopteridaceae is sister to the remaining species, which are split into two clades. 
One clade consists of Rhacidosoraceae, Diplaziopsidaceae, Desmophlebiaceae, Hemidictyaceae and Aspleniaceae.
The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and Athyriaceae.

@@

```{r divergence-times-stats}
# Count number of tips in clades by taxonomy for double checking
n_tips_in_clades <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(species, family, order) %>%
	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
	group_by(taxon) %>%
	summarize(number_tips_by_tax = n())

tree_height <- max(phytools::nodeHeights(plastid_tree_dated))

# Define function for calculating age based on two genera
get_age_from_genus_pair <- function(tip_pair) {
	sanger_sampling %>%
	filter(genus %in% tip_pair) %>%
	group_by(genus) %>%
	slice(1) %>%
	pull(species) %>%
  ape::getMRCA(plastid_tree_dated, .) %>%
	phytools::nodeheight(plastid_tree_dated, .) %>%
	magrittr::subtract(tree_height, .) %>%
	number(accuracy = 0.1)
}

# Calculate age of various clades
fern_age <- get_age_from_genus_pair(c("Equisetum", "Polypodium"))
lepto_age <- get_age_from_genus_pair(c("Osmunda", "Polypodium"))
polypodiales_age <- get_age_from_genus_pair(c("Lindsaea", "Polypodium"))
eupoly_i_age <- get_age_from_genus_pair(c("Hypodematium", "Polypodium"))
eupoly_ii_age <- get_age_from_genus_pair(c("Cystopteris", "Asplenium"))

# Get ages for large clades
clade_ages <-
	sanger_sampling %>%
	select(family, order) %>%
	unique() %>%
	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
	filter(!is.na(taxon)) %>%
	unique() %>%
	left_join(
		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
		by = "taxon") %>%
	mutate(across(c(mrca, number_tips), parse_number)) %>%
	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
	filter(monophyly == "Yes") %>%
	mutate(
		# Height is distance above root
		height = map_dbl(mrca, ~phytools::nodeheight(
			tree = plastid_tree_dated, node = .))
	) %>%
	mutate(
		# Age is the total length of the tree minus height
		age = tree_height - height
	) %>%
	# Double check that n_tips is as expected
	# (so the correct node is being used)
	left_join(n_tips_in_clades, by = "taxon") %>%
	verify(all(number_tips == number_tips_by_tax)) %>%
	select(-number_tips_by_tax)
```

We estimate the crown age of ferns to be `r fern_age` ma.
Ages of other major crown groups are leptosporangiates (`r lepto_age` ma), Polypodiales (`r polypodiales_age` ma), eupolypods I (`r eupoly_i_age` ma), and eupolypods II (`r eupoly_ii_age` ma).
Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
Ages of fern families differed from previous studies, with some families estimated as having older crown ages while some had younger (`r figure("divtimes")`).

@@

## Revisiting fossil calibration points in ferns

Although we recovered slightly older ages for ferns and other large clades relative to the most recent globally sampled fern phylogeny [@Testo2016a], our estimated ages for families do not show a clear pattern, with some older and some younger than other studies (`r figure("divtimes")`).
This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies.
Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
Therefore, it is perhaps not surprising that we obtained different estimates for the age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).

<<FIXME: add discussion of specific groups once we have final ages>>

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		plot_group = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ plot_group)
	)
```

@@

genbank_plot <-
	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
	geom_line(aes(color = type)) +
	scale_color_manual(values = compartment_cols, guide = "none") +
	# Annotate panel border manually, since the automatic border gets
	# printed on top of label annotations with theme_bw()
	annotate(
		"rect", 
		xmin = min(gb_species_by_year$year) - year_buffer,
		xmax = max(gb_species_by_year$year) + year_buffer,
		ymin = min(gb_species_by_year$n_species) - sp_buffer,
		ymax = max(gb_species_by_year$n_species) + sp_buffer,
		fill = "transparent",
		color = "grey20") +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("nuclear", "mitochondrial")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
		# Do not repel from top or bottom edges.
		ylim = c(-Inf, Inf)
	) +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("total", "plastid")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
		ylim = c(-Inf, Inf),
		min.segment.length = 0
	) +
	geom_point(
		data = fern_tree_sampling,
		aes(shape = source)
	) +
	scale_shape(
		limits = fern_tree_sampling$source, 
		labels = fern_tree_sampling$label
	) +
	scale_x_continuous(expand = expansion(0, 0)) + 
	scale_y_continuous(expand = expansion(0, 0)) + 
	labs(shape = "Study", y = "No. species", x = "Year") +
	coord_cartesian(clip = "off") +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		legend.position = "bottom",
		legend.title = element_blank(),
		plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")
	)
```

@@

```{r backbone-make}
# Make exemplar set of species in backbone representing families
family_exemp <-
	backbone_taxonomy %>%
	# Drop outgroups %>%
	filter(family != "og") %>%
	# Slice to one exemplar species per family
	group_by(family) %>%
	slice(1) %>%
	ungroup() %>%
	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
	arrange(phy_order)

# Trim tree to exemplar species, rename by family
phy <- keep.tip(plastome_tree_rooted, family_exemp$species)
new_tip_label <-
	tibble(species = phy$tip.label) %>%
	left_join(family_exemp, by = "species") %>%
	pull(family)
phy$tip.label <- new_tip_label
# Convert BS support values to numeric
phy$node.label <- parse_number(phy$node.label)
# Scale to total height of 100 for easy plotting
phy <- rescale_tree(phy, 100)

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
clade_ages %>%
	filter(level == "family") %>%
	select(family = taxon, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -family) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median"))

# Convert family errors to almost long format
family_ages_errors <-
	clade_ages %>%
	filter(level == "family") %>%
	select(family = taxon, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	select(family, matches("ts|rothfels")) %>%
	select(-matches("median"))

family_ages_errors <-
	bind_rows(
		transmute(family_ages_errors, family,
					 high = ts_high, low = ts_low, author = "ts"),
		transmute(family_ages_errors, family,
					 high = rothfels_high, low = rothfels_low, author = "rothfels")
	)

# Define shapes and colors for points
ages_shapes <- c(
	current = 16, # circle
	rothfels = 17, # triangle
	schuettpelz = 15, # square
	ts = 18 # diamond
)

ages_colors <- c(
	current = palette_okabe_ito(order = "vermillion"),
	rothfels = palette_okabe_ito(order = "bluishgreen"),
	schuettpelz = palette_okabe_ito(order = "blue"),
	ts = palette_okabe_ito(order = "orange")
)

# Format names of papers for printing
ages_sampling <-
	tribble(
		~source, ~year, ~label,
		"schuettpelz", 2009, "Schuettpelz & Pryer 2009",
		"rothfels", 2015, "Rothfels et al. 2015",
		"ts", 2016, "Testo and Sundue 2016",
		"current", 2021, "This study",
	) %>%
	mutate(source = fct_reorder(source, year))

div_times_plot <-
ggplot() +
	geom_errorbarh(
		data = family_ages_errors,
		aes(xmax = high, xmin = low, color = author, y = family),
		position = position_dodge(width = 0.5)
	) +
	geom_point(
		data = filter(family_ages_median, age_type %in% c("best", "median")),
		aes(x = age, y = family, color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	labs(x = "Age (millions of years)") +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```

```{r div-times-save, eval = params$doc_type == "doc"}
# Save plot if rendering to doc
ggsave(
	result_file(figure("divtimes"), "pdf"),
	div_times_plot, width = 7, height = 8, units = "in")
```

```{r div-times-print, fig.width = 7, fig.height = 9, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
div_times_plot
```

**`r figure("divtimes")`**. Crown age of fern families (mya) estimated by selected studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
